### 2021-10-28 documents/sections/extensions.tex
incorporated uncertainty about 'to go down'

Apart from these \PTir verbs with \rc{w(ɨ)-}, there are two irregularly inflected \gl{s_a_} verbs in \trio, \obj{ɨhtə} \qu{to go down} and \obj{weka}/\obj{oeka} \qu{to defecate} \exref{downshit}.
They have \gl{1}\gl{s_a_} markers \obj{p-} and \obj{k-}, which are otherwise entirely unattested in \trio.\footnote{Although both elements also occur in other irregular forms of these verbs \parencite[180, 325, 331]{triomeira1999}.}
At least for \qu{to go down}, the \akuriyo cognate suggests that the irregular first person form \rc{p-ɨhtə-} can be reconstructed to \PTir.
Whatever their origins, they were not affected by the extension of \rc{t-}.

\ex<downshit> Idiosyncratic \gl{1}\gl{s_a_} prefixes \parencites[294]{triomeira1999}[84]{gildea1994akuriyo}\\
\begin{tabular}[t]{@{}lll@{}}
& \trio & \akuriyo\\
\qu{go down} & \obj{p-ɨhtə-} & \obj{p-ɨtə-}\\
\qu{defecate} & \obj{k-oeka-} & ?\\
\end{tabular}
\xe

Finally, \textcite{gildea1994akuriyo} also recorded four more \akuriyo verbs seemingly not affected by innovative \rc{t-} \exref{akumov.aku}, all \obj{e}-initial movement verbs.
We have only found a \trio cognate for \obj{erama} \qu{to return}, which behaves like a regular \gl{s_a_} verb in taking \obj{s-} \exref{akumov.tri}.
Further, these verbs are not mentioned by \textcite{meira1998proto}, who was also working with \posscite{gildea1994akuriyo} data.
Given that this data potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998} and the lack of support by Meira's data, we cannot reconstruct these verbs as not being affected by the extension of \PTir \rc{t-}.

@@

@@

It has been speculated that the \trio \gl{s_a_} verb with a first person \obj{k-} marker, \obj{weka}/\obj{oeka} \qu{to defecate} might have something to do with \akuriyo \obj{k-}, which would potentially make the innovation of \rc{k-} a \PTir matter \parencite[116]{meira1998proto}.
However, this hypothesis faces a problem in that the first person form for \akuriyo \qu{to defecate} is \obj{j-ereina-} \parencite[88]{gildea1994akuriyo}.
Another hypothesis is that it originated in the corresponding \kaxui form \obj{ku-weka-} (form provided by Spike Gildea, p.c.), although the occurrence of \obj{o} in \trio would still need explanation.
%Rather, it seems likely that this irregular first person form originated in a class switch from \gl{s_p_} to \gl{s_a_} (\Cref{sec:resistant_verbs}), although it is unclear how exactly that switch resulted in the irregular form.

>>>>

%Apart from these \PTir verbs with \rc{w(ɨ)-}, there are two irregularly inflected \gl{s_a_} verbs in \trio, \obj{ɨhtə} \qu{to go down} and \obj{weka}/\obj{oeka} \qu{to defecate} \exref{downshit}.
%They have \gl{1}\gl{s_a_} markers \obj{p-} and \obj{k-}, which are otherwise entirely unattested in \trio.\footnote{Although both elements also occur in other irregular forms of these verbs \parencite[180, 325, 331]{triomeira1999}.}
%At least for \qu{to go down}, the \akuriyo cognate suggests that the irregular first person form \rc{p-ɨhtə-} can be reconstructed to \PTir.
%Whatever their origins, they were not affected by the extension of \rc{t-}.
%
%\ex<downshit> Idiosyncratic \gl{1}\gl{s_a_} prefixes \parencites[294]{triomeira1999}[84]{gildea1994akuriyo}\\
%\begin{tabular}[t]{@{}lll@{}}
%& \trio & \akuriyo\\
%\qu{go down} & \obj{p-ɨhtə-} & \obj{p-ɨtə-}\\
%\qu{defecate} & \obj{k-oeka-} & ?\\
%\end{tabular}
%\xe

In addition, \textcite{gildea1994akuriyo} recorded four more \akuriyo verbs seemingly not affected by innovative \rc{t-} \exref{akumov.aku}, all \obj{e}-initial movement verbs.
We have only found a \trio cognate for \obj{erama} \qu{to return}, which behaves like a regular \gl{s_a_} verb in taking \obj{s-} \exref{akumov.tri}.
Further, these verbs are not mentioned by \textcite{meira1998proto}, who was also working with \posscite{gildea1994akuriyo} data.
Given that this data potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998} and the lack of support by Meira's data, we cannot reconstruct these verbs as not being affected by the extension of \PTir \rc{t-}.

@@

In addition, there is an \gl{s_a_} verb \obj{ɨ(h)tə} \qu{to go down}, which has an irregular first person marker \obj{p-} in \akuriyo, also reconstructible to \PTir \exref{tridown}.
It was not affected by the extension of \akuriyo \obj{k-}, but its relative chronology to the extension of \PTir \rc{t-} is unclear (see also \cref{sec:godown}).

\ex<tridown> Tiriyoan \qu{to go down} \parencites[294]{triomeira1999}[84]{gildea1994akuriyo}\\
\begin{tabular}[t]{@{}ll@{}}
\trio & \obj{p-ɨhtə-} \\
\akuriyo & \obj{p-ɨtə-}\\
\end{tabular}
\xe

@@

%It has been speculated that the \trio \gl{s_a_} verb with a first person \obj{k-} marker, \obj{weka}/\obj{oeka} \qu{to defecate} might have something to do with \akuriyo \obj{k-}, which would potentially make the innovation of \rc{k-} a \PTir matter \parencite[116]{meira1998proto}.
%However, this hypothesis faces a problem in that the first person form for \akuriyo \qu{to defecate} is \obj{j-ereina-} \parencite[88]{gildea1994akuriyo}.
%Another hypothesis is that it originated in the corresponding \kaxui form \obj{ku-weka-} (form provided by Spike Gildea, p.c.), although the occurrence of \obj{o} in \trio would still need explanation.
%Rather, it seems likely that this irregular first person form originated in a class switch from \gl{s_p_} to \gl{s_a_} (\Cref{sec:resistant_verbs}), although it is unclear how exactly that switch resulted in the irregular form.

### 2021-10-28 documents/sections/verbs.tex
incorporated uncertainty about 'to go down'

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{t-} in \PTir \pcref{sec:taranoan} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
Rather, an irregular first person form \rc{p-ɨptə} can be reconstructed for \PTir.
While it originally resisted the extension of \rc{k-} in \PPek, \bakairi subsequently introduced it \pcref{sec:pekodian}.
It was also affected by the extension of \obj{j-} in \carijo \exref{car-32}.

\ex<car-32>\carijo \perscommpar{David Felipe Guerrero}\\
\begingl
\gla irə wat͡ʃinakano tae j-ehɨtə-e//
\glb \gl{inan}.\gl{ana} body.of.water along.bounded \gl{1}-go.down-\gl{npst}//
\glft \qu{…I go down through that guachinacán.}//
\endgl
\xe
%
While it seems that \yukpa \obj{ewu(h)tu} \qu{to go down} \parencite{meira2003primeras} is cognate, it was affected by the extension of \obj{j-}.
%todo maybe insert a yukpa example here? the form is just jewuhtu

@@

A comparative view \pcref{tab:godown} shows that while a form \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about the class of this verb.
The reflexes are fairly evenly split between \gl{s_a_} and \gl{s_p_} -- in those languages that preserve the split-\gl{s} system.
In one language, \wayana, the verb shows traits of both classes, leading us to consider it a \dbqu{mixed} verb in the synchronic analysis of \wayana.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptəka} \parencite[255]{wayanatavares2005}, containing a reflex of the transitivizer \rc{-ka}, which was restricted to \gl{s_p_} verbs in \PC \parencite{gildea2019overview}.
This leads us to posit the hypothesis that the verb may have been a normal member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class -- fully so in the languages where it is \gl{s_a_}.

\wayana-external comparative evidence supports this hypothesis:
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:descend_cog} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the (rare) transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms \parencite[263]{triomeira1999}.
Thus, it seems that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.

@@

\subsection{\trio \obj{weka}/\obj{oeka} \qu{to defecate}}
\label{sec:shit}
The comparative picture for the resistant \trio \gl{s_a_} verb \qu{to defecate} \pcref{tab:defecate} shows that \rc{weka} was clearly originally an \gl{s_p_} verb.
As in the case of \qu{to go down} \pcref{sec:godown}, this verb has a mixed \gl{s_a_}/\gl{s_p_} status in \wayana.
The \bakairi reflex \obj{əeke} seems to contain a reflex of \rc{ət-}, which would explain its status as an \gl{s_a_} verb.
While the class membership of \panare \obj{aiʔka} and \obj{iʔka} \qu{to defecate} is unknown, the former seems to contain a reflex of the detransitivizer, while the latter only consists of the root, meaning that one would assume \gl{s_a_} and \gl{s_p_} membership, respectively.

\input{floats/defecate}
%
%\subsection{The \akuriyo movement verbs}
%\label{sec:movement}
%As discussed in \cref{sec:akuriyo}, four \akuriyo \gl{s_a_} verbs are attested as not having the first person marker \obj{t͡ʃ-} in \textcite{gildea1994akuriyo}; instead, they have \obj{w-} or its phonologically reduced form ∅.
%These verbs share two properties: they are all \obj{e-}initial, and they are all movement verbs.
%However, the same dataset also contains \obj{e-}initial movement verbs with a first person marker \obj{t͡ʃ-} \exref{aku-106}.
%As for the etymology of these four verbs, we can only speak to \obj{erama} \qu{to return}, which is a detransitivized form of \obj{rama} \qu{to put back}, a pair which is also found in other Cariban languages.
%For the other three verbs, we have not encountered potential cognates, but given their \obj{e-}initial nature and their semantics, it seems plausible that they, too, are derived from transitive verbs with meanings like \qu{to put up} etc.
%In any case, \akuriyo is only sparsely documented, and there are many open questions about it; the \akuriyo data in \textcite{gildea1994akuriyo} are not fully reliable.
%
%\ex<aku-106>\akuriyo \parencite[][84]{gildea1994akuriyo}\\
%\begingl
%\gla t͡ʃ-ewai//
%\glb \gl{1}-sit.down//
%\glft \qu{I sat down.}//
%\endgl
%\xe

>>>>

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{k-} in \PPek \pcref{sec:pekodian} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
The resistance against the former extension was overcome in \bakairi.
In \akuriyo, the verb already had an irregularly inflected first person form with \obj{p-}, inherited from \PTir.

At first sight, it may seem that it was affected by the independent extensions of \obj{j-} in \carijo \exref{gowhere.car-18} and \yukpa \exref{gowhere.yuk-14}.

\pex<gowhere>
\a<car-18> \carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
However, a look at the comparative picture suggests a much more complicated situation.
\cref{tab:godown} shows all attested cognates, some already discussed, including verb class membership and aligned segments.
While a form \rc{ɨpɨtə} can be reconstructed to \PC, different (proto-)languages do not agree about the class of this verb.
The reflexes are fairly evenly split between \gl{s_a_} and \gl{s_p_} -- in those languages that preserve the split-\gl{s} system.

In one language, \wayana, the verb shows traits of both classes, leading us to consider it a \dbqu{mixed} verb in the synchronic analysis of \wayana.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<gowhere>
\a<car-18> \carijo \parencite[][102]{guerrero2019carijo}\\
\begingl
\gla əji-wa-e j-eh-ɨ//
\glb \gl{2}-search-\gl{sup} \gl{1}-come-\gl{pfv}//
\glft \qu{I came looking for you.}//
\endgl
\a<yuk-14> \yukpa \parencite[][]{meira2003primeras}\\
\begingl
\glpreamble aw yéwtu//
\gla aw j-ewuhtu//
\glb \gl{1}\gl{pro} \gl{1}-go.down//
\glft \qu{I went down.}//
\endgl
\xe
%
Its causativized form is \obj{ɨptəka} \parencite[255]{wayanatavares2005}, containing a reflex of the transitivizer \rc{-ka}, which was restricted to \gl{s_p_} verbs in \PC \parencite{gildea2019overview}.
This leads us to posit the hypothesis that the verb was a regular member of the \gl{s_p_} class in pre-\wayana, but partially switched to the \gl{s_a_} class.
This in turn implies that all reflexes of this verb with \gl{s_a_} status were members of the \gl{s_p_} class at some point, but switched classes.

@@

\wayana-external comparative evidence supports this hypothesis:
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:godown} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the (rare) transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms \parencite[263]{triomeira1999}.
Thus, it seems that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.

These circumstances make it impossible to answer the question of whether it was affected by the extensions in \PTir, \carijo, and \yukpa.
For \PTir, we cannot establish a relative chronology of the verb class change, the introduction of the idiosyncratic marker \rc{p-}, and the extension of \rc{t-}.
It could have been an \gl{s_p_} verb at the time of the extension, it could have been an \gl{s_a_} verb and remained unaffected, or it could have been affected, but with a subsequent erasure of the results by the introduction of \rc{p-}.
For \carijo and \yukpa, we cannot know whether the verb switched verb class before the breakdown of the entire split-\gl{s} system -- this is a plausible scenario especially in the former case.
In all three cases, it is possible that the verb had \gl{s_a_} status at the time of the extension, but it is also possible that it was not even a potential target due to its \gl{s_p_} status.
On the other hand, the class change must have happened before the split-up of \trio and \akuriyo, and therefore this verb resisted the extension of \akuriyo \obj{k-}.
Likewise, the class change must have taken place before the extension of \PPek \rc{k-}, since other scenarios would require an \gl{s_p_} verb taking on regular \gl{s_a_} markers, but selecting the archaic first person marker.

@@

%\subsection{\trio \obj{weka}/\obj{oeka} \qu{to defecate}}
%\label{sec:shit}
%The comparative picture for the resistant \trio \gl{s_a_} verb \qu{to defecate} \pcref{tab:defecate} shows that \rc{weka} was clearly originally an \gl{s_p_} verb.
%As in the case of \qu{to go down} \pcref{sec:godown}, this verb has a mixed \gl{s_a_}/\gl{s_p_} status in \wayana.
%The \bakairi reflex \obj{əeke} seems to contain a reflex of \rc{ət-}, which would explain its status as an \gl{s_a_} verb.
%While the class membership of \panare \obj{aiʔka} and \obj{iʔka} \qu{to defecate} is unknown, the former seems to contain a reflex of the detransitivizer, while the latter only consists of the root, meaning that one would assume \gl{s_a_} and \gl{s_p_} membership, respectively.
%
%\input{floats/defecate}
%
%\subsection{The \akuriyo movement verbs}
%\label{sec:movement}
%As discussed in \cref{sec:akuriyo}, four \akuriyo \gl{s_a_} verbs are attested as not having the first person marker \obj{t͡ʃ-} in \textcite{gildea1994akuriyo}; instead, they have \obj{w-} or its phonologically reduced form ∅.
%These verbs share two properties: they are all \obj{e-}initial, and they are all movement verbs.
%However, the same dataset also contains \obj{e-}initial movement verbs with a first person marker \obj{t͡ʃ-} \exref{aku-106}.
%As for the etymology of these four verbs, we can only speak to \obj{erama} \qu{to return}, which is a detransitivized form of \obj{rama} \qu{to put back}, a pair which is also found in other Cariban languages.
%For the other three verbs, we have not encountered potential cognates, but given their \obj{e-}initial nature and their semantics, it seems plausible that they, too, are derived from transitive verbs with meanings like \qu{to put up} etc.
%In any case, \akuriyo is only sparsely documented, and there are many open questions about it; the \akuriyo data in \textcite{gildea1994akuriyo} are not fully reliable.
%
%\ex<aku-106>\akuriyo \parencite[][84]{gildea1994akuriyo}\\
%\begingl
%\gla t͡ʃ-ewai//
%\glb \gl{1}-sit.down//
%\glft \qu{I sat down.}//
%\endgl
%\xe
